Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-25 paper/paper.tex
paper: New corrections
Co-Authored-By: Cesar Guzmann <15159632+cguz@users.noreply.github.com>

\section{Open Cluster Characterization method}
\label{sec:aims_and_method}

@@

\subsection{Selection of the catalog}

@@

\begin{itemize}
  \item \textbf{deep autoencoder:} it is a neural network that learns to copy its input
        to its output. It is composed of encoder layers followed by decoder layers.
        We train it before passing the data through the next component, the
        \emph{clustering layer}. Recent research~\cite{vincent2010stacked, hinton2006reducing}
         showed that the autoencoder provides meaningful and well-separated representations
         on real-world datasets.
  \item \textbf{clustering layer:} it receives data transformed by the autoencoder
        and iteratively is trained until a convergence criterion is met.
\end{itemize}

@@

As we can see in the parallax distribution, lower values of the number of clusters (e.g., N=2) assign
many stars to the group \emph{0} (purple) that should belong to group \emph{g1}. These badly assigned stars
are the ones that have the parallax far from the 7.3mas resonance. In contrast, large values of the
number of clusters allow us to isolate more accurately the resonance in parallax at \(\approx 7.3mas\).
For instance, at N=8, more groups are better compartmentalized. Group \emph{g2} would be the one corresponding
to the OC that we are looking for, and all the other groups are those that we would like to be a single
group. That is the stars that do not belong to the OC. The disadvantage is that K-Means with N=8 forms more
groups of stars, which complicates the task of finding the desired OC, since we would like to get just two
groups. Therefore, with a K-Means we have to find  a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups. To solve this issue, we can try to
estimate the best number of clusters by using the \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

%% Comparación con K-Means, DEC y DEC filter
\begin{figure*}[!htbp]
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_hr_diagram_melotte_22.png}
      \caption{K-Means}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.png}
      \caption{DEC}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_filtered_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_filtered_melotte_22.png}
      \caption{DEC (filt.)}
  \end{subfigure}
  \caption{Evolution of Melotte 22 characterization}
  \label{fig:melotte_22_characterization_evolution}
\end{figure*}

We have tested our model with the Melotte 25 dataset, which is one of the most difficult dataset
to characterize OC by its distribution and extension. In order to support the discussion presented
in Section~\ref{sec:deep_open_clustering}, we compared our model against a K-Means classification
algorithm. In addition, we perform a second execution only in our model where we filtered the
quantiles in the Melotte 22 dataset.

@@

Figure~\ref{fig:melotte_22_characterization_evolution} shows the evolution
of the characterization of Melotte 22. The first column corresponds to the
characterization obtained by applying K-Means alone to the dataset. The middle
column is the characterization made by the DEC model. And the column on the right
shows the characterization made by the DEC and having filtered the quantiles
lower than 0.10 and higher than 0.90. The top row shows the proper motion
configuration space while the bottom one shows the Hertzsprung–Russell diagrams.

@@

Melotte 25 is located at 66.725 degrees in right ascension and 15.867
degrees in declination with a radius of 165 arcmin.
This region contains more than 400,000 stars.

Table~\ref{tab:results_melotte_25} shows a results summary for Melotte 25.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{\(\mu_{\alpha}\) \((mas \cdot yr^{-1})\)} & \emph{\(\mu_{\delta}\) \((mas \cdot yr^{-1})\)}
        & \emph{\( \varpi \) \((mas)\)} & \emph{\# stars} \\
        \hline
        \textbf{Simbad}\tablefootnote{Results have been taken from the SIMBAD astronomical database\cite{wenger2000simbad}}
        & 104.92 \( \pm \) 0.12 & -28.00 \( \pm \) 0.09 & 21.052 \( \pm \) 0.065 & - \\
        Clusterix & 106.796 \( \pm \) 6.229 & -24.870 \( \pm \) 5.417 & 21.210 \( \pm \) 1.115 & 109 \\
        K-Means & 79.936 \( \pm \) 3.7 & -45.362 \( \pm \) 3.8 & 20.901 \( \pm \) 0.31 & 374 \\
        DEC & 104.051 \( \pm \) 3.2 & -33.424 \( \pm \) 2.5 & 22.072 \( \pm \) 0.13 & 219 \\
        \textbf{DEC (filt.)} & 105.96 \( \pm \) 3.5 & -30.00 \( \pm \) 2.4 & 21.74 \( \pm \) 0.07 & 175 \\
      \end{tabular}
    }
    \caption{Melotte 25 results.}
    \label{tab:results_melotte_25}
  \end{center}
\end{table}

Parameters shown are proper motion in right ascension and declination, parallax
with their respective deviations and number of stars corresponding to the OC.

Figure~\ref{fig:result_melotte_25_dec} shows the groups found using the DEC model.
This model labels Melotte 25 as group \emph{g5}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.30\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_pm_melotte_25.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.30\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_parallax_melotte_25.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.30\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_hr_diagram_melotte_25.png}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 25 characterization using the DEC model.
           From the left to the right: proper motion configuration
           space, parallax histogram and H-R diagram.}
  \label{fig:result_melotte_25_dec}
\end{figure}

@@

\begin{table}[h]
  \begin{center}
    \begin{tabular}{l|c}
      \textbf{Hyperparameter} & \textbf{Value} \\
      \hline
      Number of Clusters & 9 \\
      Clustering Layer & \(\left[ 50, 50, 40 \right]\) \\
      Kernel Initializer Seed & 11 \\
      Quantil Threshold & 0.1 \\
    \end{tabular}
    \caption{Melotte 25 DEC hyperparameters.}
    \label{tab:hyperparameters_melotte_25}
  \end{center}
\end{table}

\section{Result discussions}

>>>>

\section{Open Cluster Characterization Method}
\label{sec:aims_and_method}

@@

\subsection{Selection of the Catalog}

@@

\begin{itemize}
  \item \textbf{deep autoencoder:} it is a neural network that learns to copy its input
        to its output. It is composed of encoder layers followed by decoder layers.
        We trained it before passing the data through the next component, the
        \emph{clustering layer}. Recent research~\cite{vincent2010stacked,hinton2006reducing}
        showed that the autoencoder provides meaningful and well-separated representations
        on real-world datasets.
  \item \textbf{clustering layer:} it receives data transformed by the autoencoder
        and iteratively is trained until a convergence criterion is met.
\end{itemize}

@@

As we can see in the parallax distribution, lower values of the number of clusters (e.g., N=2) assign
many stars to the group \emph{g0} (purple) that should belong to group \emph{g1}. These badly assigned stars
are the ones that have the parallax far from the 7.3mas resonance. In contrast, large values of the
number of clusters allow us to isolate more accurately the resonance in parallax at \(\approx 7.3mas\).
For instance, at N=8, more groups are better compartmentalized. Group \emph{g2} would be the one corresponding
to the OC that we are looking for, and all the other groups are those that we would like to be a single
group. That is the stars that do not belong to the OC. The disadvantage is that K-Means with N=8 forms more
groups of stars, which complicates the task of finding the desired OC, since we would like to get just two
groups. Therefore, with a K-Means we have to find a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups. To solve this issue, we can try to
estimate the best number of clusters by using the \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

%% Comparación con K-Means, DEC y DEC filter
\begin{figure*}[!htbp]
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/kmeans_pm_melotte_25.png}
      \includegraphics[width=\textwidth]{../figures/melotte_25/kmeans_hr_diagram_melotte_25.png}
      \caption{K-Means}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_pm_melotte_25.png}
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_hr_diagram_melotte_25.png}
      \caption{DEC}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_pm_filtered_melotte_25.png}
      \includegraphics[width=\textwidth]{../figures/melotte_25/dec_hr_diagram_filtered_melotte_25.png}
      \caption{DEC (filt.)}
  \end{subfigure}
  \caption{Evolution of Melotte 25 characterization}
  \label{fig:melotte_25_characterization_evolution}
\end{figure*}

We have tested our model with the Melotte 25 dataset, which is one of the most difficult dataset
to characterize OC by its distribution and extension, because it is located in a region with
400,000 stars.

In order to support the discussion presented in Section~\ref{sec:deep_open_clustering}, we compared
our model against a K-Means classification algorithm. In addition, we perform a second execution
only in our model where we filtered the quantiles in the Melotte 25 dataset.

@@

Figure~\ref{fig:melotte_25_characterization_evolution} shows the evolution of the characterization
of Melotte 25. The first column corresponds to the characterization obtained by applying K-Means
(with N=10) alone to the dataset. The middle column is the characterization made by our model.
And the column on the right shows the characterization made by our model but filtering the
quantiles lower than 0.10 and higher than 0.90. The top row shows the proper motion configuration
space, while the bottom one shows the Hertzsprung–Russell diagrams.

@@

The clustering made by K-Means arises 9 groups, but most of them contain stars that do not belong
to the Melotte 25 OC. However, after applying our model, most of the stars are grouped into a single
cluster. One of the other groups is the OC we are looking for, and the other seven could be new OCs
inside the studied region.

The dataset Melotte 25 is located at 66.725 degrees in right ascension and 15.867 degrees in
declination with a radius of 165 arcmin. As it was mentioned before, this region contains more than
400,000 stars making it difficult to group stars of the same region.

Table~\ref{tab:results_melotte_25} summarizes the results for the Melotte 25 dataset. The first column
shows the name of the method. Second and third columns describe the proper motion in right ascension and
declination, respectively. The fourth column shows the parallax. And the last column presents the number
of stars corresponding to the OC. Columns from second to fourth have their respective deviations computed
as the standard error of the predicted group.

By comparing the results obtained with our model, we can see that they are in range with those extracted
from SIMBAD astronomical database~\cite{wenger2000simbad}. While the results provided by K-Means model are
far from the official data. This comparison demonstrates that our model improves the results obtained with
a raw K-Means.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{\(\mu_{\alpha}\) \((mas \cdot yr^{-1})\)} & \emph{\(\mu_{\delta}\) \((mas \cdot yr^{-1})\)}
        & \emph{\( \varpi \) \((mas)\)} & \emph{\# stars} \\
        \hline
        \textbf{Simbad}\tablefootnote{Results have been taken from the SIMBAD astronomical database~\cite{wenger2000simbad}}
        & 104.92 \( \pm \) 0.12 & -28.00 \( \pm \) 0.09 & 21.052 \( \pm \) 0.065 & - \\
        %Clusterix & 106.796 \( \pm \) 6.229 & -24.870 \( \pm \) 5.417 & 21.210 \( \pm \) 1.115 & 109 \\
        K-Means & 79.936 \( \pm \) 3.7 & -45.362 \( \pm \) 3.8 & 20.901 \( \pm \) 0.31 & 374 \\
        DECOCC & 104.051 \( \pm \) 3.2 & -33.424 \( \pm \) 2.5 & 22.072 \( \pm \) 0.13 & 219 \\
        \textbf{DECOCC (filt.)} & 105.96 \( \pm \) 3.5 & -30.00 \( \pm \) 2.4 & 21.74 \( \pm \) 0.07 & 175 \\
      \end{tabular}
    }
    \caption{Summary for the Melotte 25 results.}
    \label{tab:results_melotte_25}
  \end{center}
\end{table}

Figure~\ref{fig:result_melotte_25_dec} shows the groups found using our model.
Our model labels Melotte 25 as group \emph{g5}.

\begin{figure*}[!hbt]
  \centering
  \begin{subfigure}[t]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_25/dec_pm_melotte_25.png}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_25/dec_parallax_melotte_25.png}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_25/dec_hr_diagram_melotte_25.png}
  \end{subfigure}
  \caption{Melotte 25 characterization.
           From left to right: proper motion configuration space,
           parallax histogram and H-R diagram.}
  \label{fig:result_melotte_25_dec}
\end{figure*}

@@

\begin{table}[h]
  \begin{center}
    \begin{tabular}{l|c}
      \textbf{Hyperparameter} & \textbf{Value} \\
      \hline
      Number of Clusters & 9 \\
      Clustering Layer & \(\left[ 50, 50, 40 \right]\) \\
      Kernel Initializer Seed & 11 \\
      Quantil Threshold & 0.1 \\
    \end{tabular}
    \caption{Hyperparameters for Melotte 25.}
    \label{tab:hyperparameters_melotte_25}
  \end{center}
\end{table}

These hyperparameters are used to fine-tuning our model and improve its accuracy. Although they depend on
the studied region, no one of them is related with physical properties of the cluster. That makes easier
to automate the selection of hyperparameters.

\paragraph{Final Discussion.}
